Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-09 Paper/paper.tex
get post processing to work

\includegraphics[width=1.\columnwidth]{plots/pdfLambda_vs_N_L800.pdf}
\includegraphics[width=1.\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L800.pdf}
\caption{{\it Left}:  Posterior probability distributions for $\lambdans$
and associated $90\%$ confidence intervals, shown as a function of number of events 
accumulated $N$.
{\it Right}: Filled-region plots showing the median and $90\%$ confidence intervals
for $\lambdans$, as a function of the number of observed events $N$.
 Shown are two independent filled regions. 

@@

\includegraphics[width=1.\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L2000.pdf}
\includegraphics[width=1.\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L1500.pdf}\\
\includegraphics[width=1.\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L1000.pdf}
\includegraphics[width=1.\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L500.pdf}

@@

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
% 
Imagine we have $N$ stretches of data,
$d_1, d_2, \cdots, d_N$, each containing a single observation of an NSBH binary. Let $K$
denote our collective prior knowledge, except for expectations on binary parameters
themselves, which will enter the following calculation explicitly. Let the gravitational wave
signal from each binary be characterized by the parameters 
$\,\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}$ and $\lambdans$. Note that we have folded
in the luminosity distance, inclination angle of the binary's orbital angular momentum with
respect to the plane of the detector, and its sky location angles into a single parameter $\rho$,
which is the SNR of the received signal. Note also that $\vec{\theta}$ does not include
$\lambdans$ as the latter would be treated independently. In addition we assume that all
NS's in a population will have the same deformability parameter $\lambdans$, and therefore
its cumulative measurement is possible.
% 
Using Bayes' theorem, the measured probability distribution function for $\lambdans$
from $N$ observations is
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{1-N}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and we have assumed
that all of the $N$ observations are independent. A priori, we assume that no 
particular value of $\lambdans$ is more preferred over another, and $\lambdans\leq 4000$,
i.e.
\begin{equation}\label{eq:lprior}
 p(\lambdans | K) = \dfrac{1}{4000}\,\mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right)
\end{equation}
% 
The second set of terms in Eq.~\ref{eq:lambdaMultiple} of the form $p(\lambdans | d_i, K)$
is each the probability distribution for $\lambdans$ inferred {\it a posteriori} from
the \textit{i}-th observation independently. 
They are calculated by integrating
\begin{equation}
 p(\lambdans | d_i, K) = \int \D \vec{\theta}\, p(\vec{\theta}, \lambdans | d_i, K),
\end{equation}
where $p(\vec{\theta}, \lambdans | d_n, K)$ is the inferred joint probability 
distribution of all source parameters $\vec{\theta}\cup\{\lambdans\}$ for the 
$i$-th event, as described in Sec.~\ref{s2:bayesian}.

Our goal is to understand the improvement in our measurement of $\lambdans$
with the number of recorded events. To do so, we simulate populations of $N$
observations each, and quantify what we learn from each successive observation 
using Eq.~\ref{eq:lambdaMultiple}. This allows us to determine how
rapidly does our median estimate for $\lambdans$ converges to the true value,
and how broad our uncertainty intervals for the same are, as $N$ increases.
In order to generate each population, we first fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ some value that is chosen from
$\{500,800,1000,1500,2000\}$. Next, we successively generate $N$ 
independent events, uniformly sampling the remaining
source parameters from the following ranges: (i) mass-ratio $q\in[2,5]$, (ii)
BH spin $\chibh\in[0, 0.75]$, (iii) effective distance from within a 
spherical volume whose radius is given by the lowest detectable
SNR (which we take as $\rho=10$). We remind ourselves that
once source parameters are fixed, there is a one-to-one mapping between
effective distance and SNR. Finally, in order to mitigate computational
cost, we make an additional simplification.
We perform full MCMC parameter estimation for a set of systems with parameters
laid on the regular hypercubic grid (henceforth ``G'') that spans
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
The final population is given by substituting all of the $N$ events, generated
as above, with their respective nearest neighbours on the grid G.

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show results for a population
of $N=50$ events with $\lambdans=800$. In the left panel, we show the 
probability distributions for $\lambdans$ inferred from $N$ events with $N$
ranging from $1-50$. We also mark the $90\%$ confidence interval on $\lambdans$
as dark grey vertical lines corresponding to each cumulative distribution.
Distributions are normalized to unit area. We see that the information coming
from the first few observations does not limit $\lambdans$ much more than our
prior from Eq.~\ref{eq:lprior} does. However, the median of the distribution 
tracks the true value of $\lambdans$ fast, reaching close within as few as $4$
observations.
% 
In the right panel, we show our measurement uncertainty $\delta\Lambda/\Lambda$
as a function of $N$, where we take $\delta\Lambda$ to be the width of the 
$90\%$ confidence interval on $\lambdans$ after $N$ observations.
This is given by the green filled region (left y-axis).
The green filled circles show the median measured value of $\lambdans$
as a function of $N$. The dashed line at y value $=1$ marks zero error, i.e.
$\delta\Lambda/\Lambda=0$. The dotted and dashed lines around this denote
$\pm 25\%$ and $\pm 50\%$ relative errors, respectively. Independently, the
light-grey filled shown the absolute measurement uncertainty $\delta\Lambda$
on the right y-axis , as a function of $N$. A horizontal grey dashed line
marks the true value of $\lambdans$. Contours of $1/\sqrt{N}$ are shown as 
dotted light grey curves to guide the eye while studying the dependence of
$\delta\Lambda$ on $N$.
%

>>>>

\includegraphics[width=1.05\columnwidth]{plots/pdfLambda_vs_N_L800.pdf}
% \includegraphics[width=1.\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L800.pdf}
\caption{%
% {\it Left}:
Posterior probability distributions for $\lambdans$ (colored curves), and 
associated $90\%$ confidence intervals (grey vertical lines), shown for 
different number of accumulated observations $N$.
% {\it Right}: Filled-region plots showing the median and $90\%$ confidence intervals
% for $\lambdans$, as a function of the number of observed events $N$.
%  Shown are two independent filled regions. 
% %  
%  One in green shows the recovered confidence interval
%  for $\lambdans$ (on the left y-axis), normalized by its true value,
%  as a function of the 
%  number of observed events $N$. The recovered median is shown by the
%  green line-circled curve. The pair of horizontal green dashed (dotted)
%  lines show the $\pm 50\%$ ($\pm 25\%$) symmetric error bounds around
%  the true $\lambdans$ value, which is shown at y-value $=1$ by definition.
% % 
%  The second filled region, in grey, shows the recovered
%  confidence interval for neutron star compactness, without normalizing.
%  The corresponding y-values are shown on the *right* y-axis. The horizontal grey
%  dashed line traces the true value of $\lambdans$, and is what we expect
%  the filled region to approach with increasing number of observations.
% %  
%  Finally, the grey dotted lines are contours of $1/\sqrt{N}$, drawn to aid the eye.

@@

\includegraphics[width=.9\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L2000.pdf}
\includegraphics[width=.9\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L1500.pdf}\\
\includegraphics[width=.67\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L1000.pdf}
\includegraphics[width=.67\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L800.pdf}
\includegraphics[width=.67\columnwidth]{plots/FillBetweenNormErrorBarsLambda_vs_N_L500.pdf}

@@

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
% 
Imagine that we have $N$ stretches of data, $d_1, d_2, \cdots, d_N$, each 
containing a single signal emitted by an NSBH binary. Each of these signals can
be characterized by the following binary parameters:
$\,\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}\cup\{\lambdans\}$. Note
that we have folded in the luminosity distance, inclination angle of the 
binary's orbital angular momentum with respect to the plane of the detector,
and source's sky location angles into a single parameter $\rho$, which is the 
SNR of the received signal. This is possible because of our simplification of 
using only the dominant $l=|m|=2$ modes of the GW strain~\footnote{The 
$l=|m|=2$ multipoles of the GW strain, when decomposed in a spin $-2$ weighted
spherical harmonic basis, contain more than $99\%$ of the total signal power}
that makes various angle parameters completely degenerate with distance. 
Further, let $K$ denote all of our collective prior knowledge, except for 
expectations on binary parameters themselves, which enter the following 
calculation explicitly. For instance, $K$ includes our assumption that all
NS's in a single population have the same deformability parameter $\lambdans$, 
and its cumulative measurement is therefore possible.
% 
Using Bayes' theorem, the measured probability distribution function for 
$\lambdans$ from $N$ observations is given by~\footnote{%
Eq.~\ref{eq:lambdaMultiple} includes the following implicit factor to normalize
its right-hand side: $\dfrac{\prod_i p(d_i|K)}{\int p(\lambdans)\,p(d_1, d_2, \cdots, d_N | \lambdans; K)\,\D\lambdans}$}
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{1-N}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and all $N$ 
observations are taken to be mutually independent. A priori, we assume
that no particular value of $\lambdans$ is preferred over another, and 
$\lambdans\leq 4000$, i.e.
\begin{equation}\label{eq:lprior}
 p(\lambdans | K) = \dfrac{1}{4000}\,\mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right).
\end{equation}
% 
In the second set of terms in Eq.~\ref{eq:lambdaMultiple} (of the form 
$p(\lambdans | d_i, K)$), each is the probability distribution for $\lambdans$
inferred {\it a posteriori} from the \textit{i}-th observation in isolation. 
They are calculated by integrating
\begin{equation}\label{eq:margpost}
 p(\lambdans | d_i, K) = \int \D \vec{\theta}\, p(\vec{\theta}, \lambdans | d_i, K),
\end{equation}
where $p(\vec{\theta}, \lambdans | d_n, K)$ is the inferred joint probability 
distribution of all source parameters $\vec{\theta}\cup\{\lambdans\}$ for the 
$i$-th event, as given by Eq.~\ref{eq:postprob}. By substituting
Eq.~\ref{eq:lprior}-\ref{eq:margpost} into Eq.~\ref{eq:lambdaMultiple}, we
calculate the probability distribution for $\lambdans$ as measured using $N$
independent events.

Our goal is to understand the improvement in our measurement of $\lambdans$
with the number of recorded events. To do so, we simulate populations of $N$
observations each, and quantify what we learn from each successive observation 
using Eq.~\ref{eq:lambdaMultiple}. For each population, we determine how
rapidly our median estimate for $\lambdans$ converges to the true value,
and how rapidly our confidence intervals for the same shrink, with increasing
$N$. In order to generate each population, we first fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ fixed value chosen from
$\{500,800,1000,1500,2000\}$. Each event is generated independently, sampling
the remaining source parameters uniformly from the following ranges: (i)
mass-ratio $q\in[2,5]$, (ii) BH spin $\chibh\in[0, 0.75]$, (iii) effective 
distance from within a spherical volume whose radius is given by the lowest 
detectable SNR (which we take as $\rho=10$). We remind ourselves that
once binary parameters are fixed, there is a one-to-one mapping between
effective distance and SNR. Finally, in order to mitigate computational
cost, we make an additional simplification.
We perform full Bayesian parameter estimation for a set of simulated signals
whose parameters are the vertices of the following regular hypercubic grid
(henceforth ``G''):
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
The final population is given by substituting all of the $N$ events, generated
as above, with their respective nearest neighbours on the parameter grid G.

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show results for a population
of $N=80$ events. The neutron star deformability is fixed at $\lambdans=800$
for all events. Each curve shows the probability distributions for $\lambdans$
as inferred from $N$ events, with $N$ ranging from $1-50$. We also mark the 
$90\%$ confidence intervals for $\lambdans$ as dark grey vertical lines,
one associated with each probability distribution curve. Distributions are 
normalized to unit area. From the figure, we can visualize the information 
information flow. The first few observations do not have enough information 
to bound $\lambdans$ (with $90\%$ confidence) much more than our prior from
Eq.~\ref{eq:lprior} does. However, the median of these cumulative distributions
track the true value of $\lambdans$ rapidly, reaching close in as few as $4$
observations.

In Fig.~\ref{fig:TT_Lambda_vs_N_L500_2000_CI90_0}, we show our measurement 
uncertainty for $\lambdans$ for different populations, with $\lambdans$ alone
changing between populations. Lets consider the top left panel first, which
corresponds to a population of the most deformable neutron stars considered,
with $\lambdans=2000$. The green filled region shows $\delta\Lambda/\Lambda$
as a function of $N$, where we take $\delta\Lambda$ to be the width of the 
$90\%$ confidence interval on $\lambdans$ after $N$ observations.
%
The green filled circles show the median measured value of $\lambdans$
as a function of $N$. The dashed line at y value $=1$ marks zero error, i.e.
$\delta\Lambda/\Lambda=0$. The dotted and dashed lines around this denote
$\pm 25\%$ and $\pm 50\%$ relative errors, respectively. Independently, the
light-grey filled shown the absolute measurement uncertainty $\delta\Lambda$
on the right y-axis , as a function of $N$. A horizontal grey dashed line
marks the true value of $\lambdans$. Contours of $1/\sqrt{N}$ are shown as 
dotted light grey curves to guide the eye while studying the dependence of
$\delta\Lambda$ on $N$.
%
